Controlling synchronization in coupled area-preserving maps by stickiness

### 2018-09-26 Syncarea_control/paper.tex
paper and ref updated

\documentclass[reprint,superscriptaddress,amsmath,amssymb,aps,pre]{revtex4-1}

@@

\section{Introduction} 
A low-dimensional Hamiltonian system commonly exhibits a mixed phase space 
i.e. regular structures and chaotic regions may co-exist at a given degree of 
nonlinearity. This mixed nature has interesting consequences for the transport
properties of such systems, for instance, the existence of anomalous kinetics,
L\'{e}vy processes and L\'{e}vy flights \cite{Klafter1994,Zaburdaev2015}, 
power law contributions to recurrence and other statistics, and the existence 
of dynamical traps \cite{Zaslavsky2002a,Zaslavsky2002b}. An interesting region 
of a mixed phase space is the interface between regular regions and chaotic 
sea \cite{Mackay1984, Easton1993}. The dynamics in these neighborhoods are 
complex but fairly well understood \cite{Meiss2015} for low dimensional 
systems. The complexity arises from the fact that a chaotic trajectory spends 
an arbitrary long but finite time at the boundaries of regular islands before 
exiting to the chaotic sea. The intermittent tendency of chaotic trajectories 
to stay close to the regular boundaries is called stickiness. A major 
consequence of the stickiness is  the existence of power law in the 
Poincar{\'e} recurrence times indicating algebraic decay for long 
times times rather than exponential decay expected for normal transport. 
Therefore, due to stickiness, even a small regular island can influence the 
global transport properties of the system and decay of correlations. The 
phenomenon has been of great interest and continues to be studied on 
theoretical level \cite{Altmann2006, Altmann2010,Livorati2012,Bunimovich2012, 
Kruger2015, Lange2016}. In addition, stickiness has found application in 
several problems such as particle advection in fluids 
\cite{Babiano1994,Tel2005}, transport in 
plasma fusion devices \cite{Szezech2012,Martins2014}, celestial mechanics 
\cite{Efthymiopoulos1999, Harsoula2010, Harsoula2016}.

@@

Synchronization in coupled chaotic systems, coupled map lattices and networks 
have been studied in details (for reviews, see 
\cite{Pecora1998,Boccaletti2002, Pecora2015}). 
Coupled Hamiltonian systems are also known to show synchronization such as 
measure synchronization~\cite{Hampton1999,Wang2003,Vincent2005,Gupta2017} and 
identical synchronization~\cite{Mahata2016,Das2017} The fact however remains 
that the ability to control synchronization based on the dynamics of the 
process has not been given much attention, until 
recently~\cite{Grabow2011,Wang2016}. In realistic systems, the speed of 
synchronization may be significant. For example, in
neurosciences,  the speed of the visual and olfactory processing are 
interesting problems \cite{Thorpe1996,Uchida2003}. In this paper, we intend 
to develop a couple of 
mechanisms to control synchronization process to wit,  to increase and to 
decrease synchronization times based on the location of synchronization traps 
in the sticky neighborhood of regular islands in the phase space of the 
area-preserving system of the standard map. We demonstrate the mechanism at $K 
= 6$ where only two small regular islands exist.

@@

\section{Delayed Synchronization Times}
\label{sec:delay}
In order to develop a numerical technique to delay synchronization, we  first 
have to identify the vicinity of regular islands in the phase space. A 
mechanism  to suppress stickiness based on the knowledge of hyperbolic and non 
hyperbolic regions in the phase has been reported recently~\cite{Kruger2015}. 
Our method, however, targets the trajectories themselves in the sticky region 
to delay the process.  It is to be noted that the proposed procedure ignores 
the details of a rather complex hierarchy of structures around regular 
islands. We employ the edge-detection algorithm due to Benkadda {\it et al.} 
\cite{Benkadda1997} to identify the edges of regular islands in phase space. 
For the numerical procedure, we divide the phase space in $100 \times 100$ 
grid. The edges of both islands are detected by applying the standard map 
to point initiated in the chaotic sea, for $10^7$ times.  The edges thus 
detected in shown in Fig~\ref{fig:edge_reg_island}.  We have chosen the 
proximity parameter $d=0.02$ i.e. Euclidean distance from points 
on the edge indicate the extent of the vicinity of the island.  This vicinity, 
therefore, indicate the domain wherein the synchronization traps exist.  To 
demonstrate this explicitly, we compare the phase angles, defined by $\theta = 
\tan^{-1}(\frac{Q}{P}$), of the points on numerically detected edges and 
the points of synchronization, as 
follows.

@@

Fig.~\ref{fig:Edge_Sync_angles}(b) shows the distribution of these phase angles for about 50 000 randomly chosen initial conditions for the drive and response maps that lead to synchronization. The locations of the peaks in the distribution shown in (a) approximately matches with this in (b). It is, therefore, visibly clear that synchronization occurs in the same angular domain of the phase space where the regular islands exist. Our numerically detected edges, thus correctly locate the synchronization traps in the phase space.  We now discuss the control mechanism to delay synchronization time.

@@

For the numerical implementation of the mechanism, we have considered $1000$ 
randomly chosen initial conditions from the chaotic sea of the phase space at 
$K = 6$. The maps have been iterated for $10^7$ times in each case.  The 
sticky domain is the determined by the proximity parameter $d_0$ around the 
numerically detected edges of the regular islands. We demonstrate the 
effectiveness of the control procedure  with respect to several $n-$step 
controls. The computations have been performed for $n \in \mathbb{A}$, where 
$\mathbb{A} = \{10, 20,30,...,90, 100, 150,200, ... ,10 000\}$.

@@

The degree of controlled synchronization $\beta$ may be defined as the ratio 
of synchronization times without numerical control, $\tau_c$ to 
synchronization times $\tau_0$ without control, i.e. $\beta =  
\ln(\frac{\tau_d}{\tau_0}$).  Therefore, for successfully delayed 
synchronization $\beta>0$ indicating $\tau_c > \tau$.The plot in 
Fig.~\ref{fig:Strength_con} shows the variation of $\beta$ i.e. $\beta_{min}$, 
$\beta_{mean}$, and $\beta_{max}$  indicating minimum, mean and maximum 
$\beta$ for considered $n-$step controls. Clearly, we are able to attain 
$\beta_{max}$ values to be in the range $5$ to $14$ while mean values remain 
$\sim2$. Therefore, we are successful in achieving 
significantly large delays in synchronization times.

@@

\section{Conclusions}
\label{sec:conclusions}
To summarize, we have developed numerical procedures to control 
synchronization in two identical standard maps coupled under the 
drive-response configuration. The procedure is based on the fact that 
synchronization in the 
system typically occurs in the sticky neighborhoods of the regular islands in 
the phase space. We have shown that a delay in synchronization can be obtained 
by kicking the trajectories slightly away from the domain temporarily. The 
efficiency of the procedure depends upon the number of steps employed in 
control. By applying the procedure to the system on merely $1\%$ of the phase 
space area (at $K = 6$), we have successfully increased synchronization times. 
The maximum number of control steps that are used is $0.01\%$ of the total of 
number of iterations used throughout the computation.  Furthermore, the 
procedure gives variety of delays. The limitations of this procedure are, one, 
the deflection may insert the drive trajectory in the regular islands which 
would lead to undesirable faster synchronization and two, a predetermined 
synchronization time in not obtainable. Furthermore, we have demonstrated a 
simple way to fasten synchronization by the parameter perturbation technique 
which rapidly drives the chaotic trajectories in the sticky neighborhood of 
regular islands. The distribution of degree of controlled synchronization 
times indicates the efficiency of the procedure. In both the numerical 
procedures, we have ignored the complex hierarchical structures which lead to 
sticky behavior of the chaotic orbits in the phase space. A possible future 
direction is to investigate the role of stickiness in measure synchronization 
in coupled Hamiltonian systems.

>>>>

\documentclass[reprint,amsmath,amssymb,aps,pre]{revtex4-1}

@@

\section{Introduction} 
A low-dimensional Hamiltonian system commonly exhibits a mixed phase space 
i.e. regular structures and chaotic regions may co-exist at a given degree of 
nonlinearity. This mixed nature has interesting consequences for the transport
properties of such systems, for instance, the existence of anomalous kinetics,
L\'{e}vy processes and L\'{e}vy flights \cite{Klafter1994,Zaburdaev2015}, 
power law contributions to recurrence and other statistics, and the existence 
of dynamical traps \cite{Zaslavsky2002a,Zaslavsky2002b}. An interesting region 
of a mixed phase space is the interface between regular regions and chaotic 
sea \cite{Mackay1984, Easton1993}. The dynamics in these neighborhoods are 
complex but fairly well understood \cite{Meiss2015} for low dimensional 
systems. The complexity arises from the fact that a chaotic trajectory spends 
an arbitrary long but finite time at the boundaries of regular islands before 
exiting to the chaotic sea. The intermittent tendency of chaotic trajectories 
to stay close to the regular boundaries is called stickiness. A major 
consequence of the stickiness is  the existence of power law in the 
Poincar{\'e} recurrence times indicating algebraic decay for long 
times rather than exponential decay expected for normal transport. 
Therefore, due to stickiness, even a small regular island can influence the 
global transport properties of the system and decay of correlations. The 
phenomenon has been of great interest and continues to be studied on 
theoretical level \cite{Altmann2006, Altmann2010,Livorati2012,Bunimovich2012, 
Kruger2015, Lange2016}. In addition, stickiness has found application in 
several problems such as particle advection in fluids 
\cite{Babiano1994,Tel2005}, transport in 
plasma fusion devices \cite{Szezech2012,Martins2014}, celestial mechanics 
\cite{Efthymiopoulos1999, Harsoula2010, Harsoula2016}.

@@

Synchronization in coupled chaotic systems, coupled map lattices and networks 
have been studied in details (for reviews, see 
\cite{Pecora1998,Boccaletti2002, Pecora2015}). 
Coupled Hamiltonian systems are also known to show synchronization such as 
measure synchronization~\cite{Hampton1999,Wang2003,Vincent2005,Gupta2017} and 
identical synchronization~\cite{Mahata2016,Das2017} However, the ability to control synchronization based on the dynamics of the process has not been given much attention, until 
recently~\cite{Grabow2011,Wang2016}. In realistic systems, the speed of 
synchronization may be significant. For example, in
neurosciences,  the speed of the visual and olfactory processing are 
interesting problems \cite{Thorpe1996,Uchida2003}. In this paper, we intend 
to develop a couple of 
mechanisms to control synchronization process to wit,  to increase and to 
decrease synchronization times based on the location of synchronization traps 
in the sticky neighborhood of regular islands in the phase space of the 
area-preserving system of the standard map. We demonstrate the mechanism at $K 
= 6$ where only two small regular islands exist.

@@

\section{Delayed Synchronization Times}
\label{sec:delay}
In order to develop a numerical technique to delay synchronization, we  first 
have to identify the vicinity of regular islands in the phase space. A 
mechanism  to suppress stickiness based on the knowledge of hyperbolic and non 
hyperbolic regions in the phase has been reported recently~\cite{Kruger2015}. 
Our method, however, targets the trajectories themselves in the sticky region 
to delay the process.  It is to be noted that the proposed procedure ignores 
the details of a rather complex hierarchy of structures around regular 
islands. We employ the edge-detection algorithm due to Benkadda {\it et al.} 
\cite{Benkadda1997} to identify the edges of regular islands in phase space. 
For the numerical procedure, we divide the phase space in $100 \times 100$ 
grid. The edges of both islands are detected by applying the standard map 
to point initiated in the chaotic sea, for $10^7$ times.  The edges thus 
detected is shown in Fig~\ref{fig:edge_reg_island}.  We have chosen the 
proximity parameter $d=0.02$ i.e. Euclidean distance from points 
on the edge indicate the extent of the vicinity of the island.  This vicinity, 
therefore, indicate the domain wherein the synchronization traps exist.  To 
demonstrate this explicitly, we compare the phase angles, defined by $\theta = 
\tan^{-1}(\frac{Q}{P}$), of the points on numerically detected edges and 
the points of synchronization, as 
follows.

@@

Fig.~\ref{fig:Edge_Sync_angles}(b) shows the distribution of these phase angles for about 50 000 randomly chosen initial conditions for the drive and response maps that lead to synchronization. The locations of the peaks in the distribution shown in ~\ref{fig:Edge_Sync_angles}(a) approximately matches with this in ~\ref{fig:Edge_Sync_angles}(b). It is, therefore, visibly clear that synchronization occurs in the same angular domain of the phase space where the regular islands exist. Our numerically detected edges, thus correctly locate the synchronization traps in the phase space.  We now discuss the control mechanism to delay synchronization time.

@@

For numerical implementation of the mechanism, we have considered $1000$ 
randomly chosen initial conditions from the chaotic sea of the phase space at 
$K = 6$. The maps have been iterated for $10^7$ times in each case.  The 
sticky domain is determined by the proximity parameter $d_0$ around the 
numerically detected edges of the regular islands. We demonstrate the 
effectiveness of the control procedure  with respect to several $n-$step 
controls. The computations have been performed for $n \in \mathbb{A}$, where 
$\mathbb{A} = \{10, 20,30,...,90, 100, 150,200, ... ,10 000\}$.

@@

The degree of controlled synchronization $\beta$ may be defined as the ratio 
of synchronization times without numerical control, $\tau_c$ to 
synchronization times $\tau_0$ without control, i.e. $\beta =  
\ln(\frac{\tau_d}{\tau_0}$).  Therefore, for successfully delayed 
synchronization $\beta>0$ indicating $\tau_c > \tau$. The plot in 
Fig.~\ref{fig:Strength_con} shows the variation of $\beta$ i.e. $\beta_{min}$, 
$\beta_{mean}$, and $\beta_{max}$  indicating minimum, mean, and maximum 
$\beta$ for considered $n-$step controls. Clearly, we are able to attain 
$\beta_{max}$ values to be in the range $5$ to $14$ while mean values remain 
$\sim2$. Therefore, we are successful in achieving 
significantly large delays in synchronization times.

@@

\section{Conclusions}
\label{sec:conclusions}
To summarize, we have developed numerical procedures to control 
synchronization in two identical standard maps coupled under the 
drive-response configuration. The procedure is based on the fact that 
synchronization in the system typically occurs in the sticky neighborhoods of the regular islands in 
the phase space. We have shown that a delay in synchronization can be obtained 
by kicking the trajectories slightly away from the domain temporarily. The 
efficiency of the procedure depends upon the number of steps employed in 
control. By applying the procedure to the system on merely $1\%$ of the phase 
space area (at $K = 6$), we have successfully increased synchronization times. 
The maximum number of control steps that are used is $0.01\%$ of the total of 
number of iterations used throughout the computation.  Furthermore, the 
procedure gives variety of delays. The limitations of this procedure are -- (a) 
the deflection may insert the drive trajectory in the regular islands which 
would lead to undesirable faster synchronization, and (b) a predetermined 
synchronization time in not obtainable. Furthermore, we have demonstrated a 
simple way to fasten synchronization by the parameter perturbation technique 
which rapidly drives the chaotic trajectories in the sticky neighborhood of 
regular islands. The distribution of degree of controlled synchronization 
times indicates the efficiency of the procedure. In both the numerical 
procedures, we have ignored the complex hierarchical structures which lead to 
sticky behavior of the chaotic orbits in the phase space. A possible future 
direction is to investigate the role of stickiness in measure synchronization 
in coupled Hamiltonian systems.
